### 2015-07-23 Harris_2015/manuscript-materials/draft.Rmd
BC_inv

***Abstract***

@@

\noindent  
***Introduction***

@@

In this paper, following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I show that Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] can provide a framework
for understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observational data. Markov networks
have been used in many scientific fields for decades in similar contexts, from physics
[where nearby particles interact magnetically with one another; @cipra_introduction_1987]
to spatial statistics [where adjacent grid cells interact with one another; @harris_contact_1974;
@gelfand_modelling_2005]. While community ecologists explored some related approaches in
the 1980's [@whittam_species_1981], computational limitations had previously imposed
severe approximations that produced unintelligible results [e.g.
"probabilities" greater than one; @gilpin_factors_1982].  Now that it is
computationally feasible to fit these models exactly, this kind of approach is
worth a second look.

@@

\noindent  ***Methods***

\noindent  
***Conditional relationships and Markov networks.***
Ecologists are often interested in learning about direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall") relationships.
The example with the shrubs and trees in
Figure 1 shows how the two measures can have opposite signs,
and suggests that conditional relationships are more relevant
for drawing inferences about species interactions (e.g. competition).

Markov networks give a probability value for every possible combination of
presences and absences in communities. For example, given a network with
binary outcomes (i.e. 0 for absence and 1 for presence), the relative
probability of observing a given presence-absence vector, $\vec{y}$, is given
by

@@

Here, $\alpha_{i}$ is an intercept term determining the amount that
the presence of species $i$ contributes
to the log-probability of $\vec{y}$; it directly controls the prevalence of
species $i$. Similarly, $\beta_{ij}$ is the amount that the co-occurrence of
species $i$ and species $j$ contributes to the log-probability; it controls
the probability that the two species will be found together (Figure 2A, Figure 2B).
$\beta$ can be interpreted as an analog of the partial covariance, but for non-Gaussian
networks [@loh_structure_2013], or as similar to a coefficient in a logistic regression model
[@lee_learning_2012]. Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends to produce assemblages
that have many pairs of positively-associated species and relatively few pairs
of negatively-associated species (exactly as an ecologist might expect).

A major benefit of Markov networks is the fact that the conditional
relationships between species can be read directly off the matrix of $\beta$
coefficients [@murphy_machine_2012]. For example, if the coefficient linking
two mutualist species is $+2$, then---all else equal---the odds of observing
either species increase by a factor of $e^2$ when its partner is present
[@murphy_machine_2012]. Of course, if all else is *not* equal (e.g. Figure 1,
where the presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from this
expectation. For this reason, it is important to consider how coefficients'
effects propagate through the network, as discussed below.

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a normalizing constant,
$Z$, which is defined so that that the probabilities of all
possible assemblages that could be produced by the model sum to one (bottom of
Figure 2B). Calculating $Z$ exactly, as is done in these analyses,
quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
required spirals exponentially into the billions and beyond.
Numerous techniques are available for working with Markov networks that keep
the computations tractable, e.g. via analytic approximations
[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
but they are beyond the scope of this paper.  Alternatively, as noted below, some
common linear and generalized linear methods can be used as computationally efficient
approximations to the full Markov network.

\noindent  
***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
however, ecologists will often need to estimate these parameters from co-occurrence data.
When the number of species is reasonably small, one can compute exact maximum
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing
$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence
frequencies (i.e. it matches the sufficient statistics of the data to the sufficient statistics of the
model distribution). For the analyses in this paper, I used the *rosalia* package
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
define the objective function and gradient as R code.  The rosalia package uses
standard convex optimization techniques to find the best values for $\alpha$ and $\beta$.

\noindent  
***Simulated landscapes.***
In order to compare different methods, I simulated two sets of landscapes using known
parameters.

The first set of simulated landscapes included the three competing species
shown in Figure 1. For each of 1000 replicates, I generated a landscape with
100 sites by sampling exactly from a probability distribution defined by the
interaction coefficients in that figure (Appendix A). Each of the methods
described below (a Markov network, two correlation-based methods and a null
model) was then evaluated on its ability to correctly infer that the two
shrub species competed with one another, despite their frequent co-occurrence.

@@

\noindent  
***Recovering species interactions from simulated data.***
I used Markov networks and  several other techniques for determining the sign and strength
of the associations between pairs of species (Appendix B).

For the Markov networks, I used the rosalia package [@harris_rosalia_2015], as described
above. Given the large number of parameters associated with some of the networks to
be estimated, I regularized the likelihood using a logistic prior distribution
with a scale of 1 on the $\alpha$ and $\beta$ terms.  This convex prior distribution has a
similar shape to the Laplace distribution used in LASSO regularization (especially in the tails),
but does not force the estimates to be exactly zero.

@@

Finally, I used the latent correlation matrix estimated by the BayesComm package [@REF]
to evaluate the claim from recent papers that the correlation coefficients estimated by “joint
species distribution models” provide an accurate assessment of species’ pairwise interactions
[@pollock_understanding_2014; see also @harris_generating_2015].

\noindent  
***Evaluating model performance.***
For the simulated landscapes based on Figure 1, method evaluation was fairly
qualitative: any method whose test statistic for the two shrubs indicated a
negative relationship passed; other methods failed. For methods that provide
confidence intervals or other inferential statistics, I reported them below.

@@

\noindent  
***Results***

\noindent  
***Three species.***
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations for 94% of the simulated
landscapes, and the randomization-based null model falsely reported such
associations 100% of the time. The two methods for evaluating conditional
relationships (Markov networks and partial covariances), however, successfully
controlled for the indirect pathway via the tree species and each correctly
identified the direct negative interaction between the shrubs 94% of the time.

\noindent  
***Twenty species.***
The accuracy of the four evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3). In general,
however, there was a consistent ordering…

The three methods that estimated marginal relationships among species and did
not explicitly control for indirect species interactions (the null model, the joint species
distribution model, and the sample correlation coefficients) differed from one another
in their $R^2$ values, but their estimates were surprisingly similar to one another.
In particular, the correlation coefficients were extremely similar to the estimates
from the other two methods (Spearman’s rho = X for the joint species distribution
model and Y for the null model).

\noindent  
***Discussion***

\noindent  
The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
a common null modeling technique reliably fails. Specifically, Markov networks
were successful even when direct interactions were largely overwhelmed by indirect
effects (Figure 1). For cases where fitting a Markov network is computationally infeasible,
these results also indicate that partial covariances and generalized linear models (the two
methods that estimated conditional relationships rather than marginal ones) can both
provide useful approximations. The partial correlations' success on simulated
data may not carry over to real data sets, however; @loh_structure_2013 show that the linear
approximations can be less reliable in cases where the true interaction matrix
contains more structure (e.g. guilds or trophic levels). Similarly, the approximation involved
in using separate generalized linear models for each species can occasionally lead to catastrophic
overfitting with small-to-moderate sample sizes [@lee_learning_2012].

@@

Despite these limitations, Markov networks have enormous potential to improve
ecological inferences. For example, Markov networks provide a simple answer
to the question of how competition should affect a species' overall
prevalence, which was a major flashpoint for the null model debates in the
1980’s [@roughgarden_competition_1983; @strong_ecological_1984]. Equation
1 can be used to calculate the expected prevalence of a species in the
absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012].
Competition's effect on prevalence in a Markov network can then be calculated
by subtracting this value and the observed prevalence (cf Figure 2D).  Apart from
BayesComm (which had the worst performance of any method evaluated here), none of
the other methods could produce this inference because they do not make predictions
about species composition. [[*back this up with more evidence or remove it*]]

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about species
interactions. Null and neutral models can be very useful for clarifying our
thinking about the numerical consequences of species' richness and abundance
patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
null model must be interpreted with care [@roughgarden_competition_1983]. In
complex networks of ecological interactions (and even in small networks with
three species), it may simply not be possible to implicate individual species
pairs or specific ecological processes like competition by rejecting a
general-purpose null [@gotelli_empirical_2009]. Estimating pairwise
coefficients directly seems like a much more promising approach: to the extent
that the models' relative performance on real data sets is similar to the
range of results shown in Figure 3, scientists in this field could easily
double their explanatory power by switching from null models to partial correlations or
generalized linear models, or triple it by switching to a Markov network. Regardless
of the specific methods ecologists ultimately choose, the most important consideration
is clearly that we need to control for indirect effects and estimate conditional relationships
between species, rather than marginal ones.

\noindent  
***Acknowledgements:***
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, N. J. Gotelli, and C. F. Dormann provided useful feedback on earlier drafts
of this work.

@@

***References:***

***Figure captions***

***Figure 1.***
**A.** A small network of three competing species. The tree (top) tends not to
co-occur with either of the two shrub species, as indicated by the strongly
negative coefficient linking them. The two shrub species also compete with one
another, as indicated by their negative coefficient (circled), but this effect
is substantially weaker. **B.** In spite of the competitive interactions
between the two shrub species, their shared tendency to occur in locations
without trees makes their occurrence vectors positively correlated (circled).
**C.** Controlling for the tree species' presence with a conditional method
such as a partial covariance or a Markov network allows us to correctly
identify the negative interaction between these two species (circled).

***Figure 2.***
**A.** A small Markov network with two species.  The depicted abiotic
environment favors the occurrence of both species ($\alpha >0$), particularly
species 2 ($\alpha_2 > \alpha_1$). The  negative $\beta$ coefficient linking
these two species implies that they co-occur less than expected under
independence.  **B.** Relative probabilities of all four possible presence-absence
combinations for Species 1 and Species 2. The exponent includes
$\alpha_1$ whenever Species 1 is present ($y_1 = 1$), but not when it is
absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
species $2$ is present ($y_2 = 1$), and $\beta$ only when both are present
($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four relative
probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** Using the
probabilities, we can find the expected frequencies of all possible co-occurrence
 patterns between the two species of interest. **D.** If $\beta$
equaled zero (e.g. if the species no longer competed for the same resources),
then the reduction in competition would allow each species to increase its
occurrence rate and the deficit of co-occurrences would be eliminated.

***Figure 3.***
Proportion of variance in interaction coefficients explained by each method
with 5, 10, or 20 species arrayed across varying numbers of sampled locations
when environmental filtering was absent (top row) or present (bottom row). A
negative $R^2$ values implies that the squared error associated with
the corresponding subset of the predictions was larger than the error one would
get from assuming that all coefficients equalled zero.

>>>>

**Abstract**

@@

\noindent  
**Introduction**

@@

In this paper, following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I show that Markov networks [undirected
graphical models also known as Markov random fields;
@murphy_machine_2012] can provide a framework for understanding
the landscape-level consequences of pairwise species interactions, and
 for detecting them from observed presence-absence matrices. Markov
networks have been used in many scientific fields for decades in similar
contexts, from physics [where nearby particles interact magnetically
with one another; @cipra_introduction_1987] to spatial statistics [where
adjacent grid cells interact with one another; @harris_contact_1974;
@gelfand_modelling_2005]. While community ecologists explored
some related approaches in the 1980's [@whittam_species_1981],
computational limitations had previously imposed severe approximations
that produced unintelligible results [e.g. "probabilities" greater than one;
@gilpin_factors_1982].  Now that it is computationally feasible to fit these
models exactly, this kind of approach is worth a second look.

@@

\noindent  **Methods**

\noindent  
**Conditional relationships and Markov networks.**
Ecologists are often interested in learning about direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall") relationships.
The example with the shrubs and trees in
Figure 1 shows how the two measures can have opposite signs,
and suggests that conditional relationships are more relevant
for drawing inferences about species interactions (e.g. competition).

Markov networks define species interactions in conditional terms, and use these
conditional relationships to determine whether a given combination of co-occurring
species is consistent with their interactions.  Given a set of binary outcomes
for each species (i.e. 0 for absence and 1 for presence), the Markov network
defines the relative probability of observing a given presence-absence vector,
$\vec{y}$, as

@@

Here, $\alpha_{i}$ is an intercept term determining the amount that
the presence of species $i$ contributes
to the log-probability of $\vec{y}$; it directly controls the prevalence of
species $i$. Similarly, $\beta_{ij}$ is the amount that the co-occurrence of
species $i$ and species $j$ contributes to the log-probability; it controls
the probability that the two species will be found together (Figure 2A, Figure 2B).
Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends to produce assemblages
that have many pairs of positively-associated species and relatively few pairs
of negatively-associated species (exactly as an ecologist might expect).

A major benefit of Markov networks is the fact that the conditional
relationships between species can be read directly off the matrix of $\beta$
coefficients [@murphy_machine_2012]. For example, if the coefficient linking
two mutualist species is $+2$, then---all else equal---the odds of observing
either species increase by a factor of $e^2$ when its partner is present
[@murphy_machine_2012]. In this way, the $\beta$ coefficients behave
very similarly to the coefficients in a logistic regression model [@lee_learning_2012].

Of course, if all else is *not* equal (e.g. Figure 1,
where the presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from this
expectation. For this reason, it is important to consider how coefficients'
effects propagate through the network, as discussed below.

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a normalizing constant,
$Z$, so that that the probabilities of all possible assemblages will sum to one (bottom of
Figure 2B). Calculating $Z$ exactly, as is done in these analyses,
quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
required spirals exponentially into the billions and beyond.
Numerous techniques are available for keeping the calculations tractable,
e.g. via analytic approximations [@lee_learning_2012] or Monte Carlo
sampling [@salakhutdinov_learning_2008], but they are beyond
the scope of this paper.  Alternatively, as noted below, some common linear
and generalized linear methods can also be used as computationally efficient
approximations to the full network.

\noindent  
**Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.**
In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
however, ecologists will often need to estimate these parameters from co-occurrence data
(i.e. from a matrix of ones and zeros indicating which species are present at which sites).
When the number of species is reasonably small, one can compute exact maximum
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients given a
presence-absence matrix by optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum represents the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence
frequencies (i.e. it matches the sufficient statistics of the model distribution to the
sufficient statistics of the data). For the analyses in this paper, I used the *rosalia* package
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
define the objective function and gradient in R code and then find the optimum.

\noindent  
**Simulated landscapes.**
In order to compare different methods, I simulated two sets of landscapes using known
parameters.

The first set of simulated landscapes included the three competing species
shown in Figure 1. For each of 1000 replicates, I generated a landscape with
100 sites by sampling exactly from a probability distribution defined by the
interaction coefficients in that figure (Appendix A). Each of the methods
described below was then evaluated on its ability to correctly infer that the two
shrub species competed with one another, despite their frequent co-occurrence.

@@

\noindent  
**Recovering species interactions from simulated data.**
I used Markov networks and  several other techniques for
determining the sign and strength of the associations between
pairs of species (Appendix B).

For the Markov networks, I used the rosalia package [@harris_rosalia_2015],
as described above. Given the large number of parameters associated with
some of the networks to  be estimated, I regularized the likelihood
using a logistic prior distribution with a scale of 2 on the $\alpha$ and
$\beta$ terms.  This convex prior distribution has a similar shape to the
Laplace distribution used in LASSO regularization (especially in the tails),
but does not force the estimates to be exactly zero.

@@

The last two methods used the latent correlation matrix estimated by the BayesComm
package [@REF] to evaluate the claim from recent papers that the correlation
coefficients estimated by “joint species distribution models” provide an accurate
assessment of species’ pairwise interactions [@pollock_understanding_2014;
see also @harris_generating_2015].  I reported both the average correlation
between each pair of species’ latent variables across 1000 Monte Carlo
samples, as well as the average partial correlation.

\noindent  
**Evaluating model performance.**
For the simulated landscapes based on Figure 1, method evaluation was fairly
qualitative: any method whose test statistic for the two shrubs indicated a
negative relationship passed; other methods failed. For methods that provide
confidence intervals or other inferential statistics (Appendix), I also reported how
often they rejected the null hypothesis of no interaction.

@@

\noindent  
**Results**

\noindent  
**Three species.**
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations for 94% of the simulated
landscapes, and the randomization-based null model falsely reported such
associations 100% of the time. The [[two]] methods for evaluating conditional
relationships (Markov networks and partial covariances), however, successfully
controlled for the indirect pathway via the tree species and each correctly
identified the direct negative interaction between the shrubs 94% of the time.

\noindent  
**Twenty species.**
The accuracy of the four evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3). In general,
however, there was a consistent ordering…

The three methods that estimated marginal relationships among species and did
not explicitly control for indirect species interactions (the null model, the joint species
distribution model’s correlation coefficients, and the sample correlation coefficients)
differed from one another in their $R^2$ values, but their estimates were surprisingly
similar to one another. In particular, the correlation coefficients were extremely similar
to the estimates  from the other two methods (Spearman’s rho = X for the joint species distribution model’s correlation matrix and Y for the null model).

\noindent  
**Discussion**

\noindent  
The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
a common null modeling technique reliably fails. Specifically, Markov networks
were successful even when direct interactions were largely overwhelmed by indirect
effects (Figure 1). For cases where fitting a Markov network is computationally
infeasible, these results also indicate that partial covariances and generalized linear
 models (the two methods that estimated conditional relationships rather than
marginal ones) can both provide useful approximations. The partial correlations'
success on simulated data may not carry over to real data sets, however;
@loh_structure_2013 show that the linear approximations can be less reliable
in cases where the true interaction matrix contains more structure (e.g. guilds
or trophic levels). Similarly, the approximation involved in using separate
generalized linear models for each species can occasionally lead to catastrophic
overfitting with small-to-moderate sample sizes [@lee_learning_2012].
For these reasons, it will usually be best to fit a Markov network
rather than one of the alternative methods when one’s computational resources
allow it.

@@

Despite these limitations, Markov networks have enormous potential to improve
ecological understanding. For example, Markov networks provide a simple answer
to the question of how competition should affect a species' overall
prevalence, which was a major flashpoint for the null model debates in the
1980’s [@roughgarden_competition_1983; @strong_ecological_1984]. Equation
1 can be used to calculate the expected prevalence of a species in the
absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012].
Competition's effect on prevalence in a Markov network can then be calculated
by subtracting this value and the observed prevalence (cf Figure 2D).  Apart from
BayesComm (which had the worst performance of any method evaluated here), none of
the other methods could produce this insight because they do not make predictions
about species composition. [[*back this up with more evidence or remove it*]]

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about species
interactions. Null and neutral models can be very useful for clarifying our
thinking about the numerical consequences of species' richness and abundance
patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
null model must be interpreted with care [@roughgarden_competition_1983]. In
complex networks of ecological interactions (and even in small networks with
three species), it may simply not be possible to implicate individual species
pairs or specific ecological processes like competition by rejecting a
general-purpose null [@gotelli_empirical_2009]. Moreover, the null model’s estimates
were extremely similar to what could be obtained from a simple correlation matrix,
which raises questions about how much additional value the null model adds.
Estimating pairwise coefficients directly seems like a much more promising approach:
to the extent that the models' relative performance on real data sets is similar to the
range of results shown in Figure 3, scientists in this field could [[easily
double their explanatory power by switching from null models to partial correlations or
generalized linear models, or triple it by switching to a Markov network]]. Regardless
of the specific methods ecologists ultimately choose, the most important consideration
is clearly that we need to control for indirect effects and estimate conditional relationships
between species, rather than marginal ones.

\noindent  
**Acknowledgements:**
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, N. J. Gotelli, and C. F. Dormann provided useful feedback on earlier drafts
of this work.

@@

**References:**

**Figure captions**

**Figure 1.**
**A.** A small network of three competing species. The tree (top) tends not to
co-occur with either of the two shrub species, as indicated by the strongly
negative coefficient linking them. The two shrub species also compete with one
another, as indicated by their negative coefficient (circled), but this effect
is substantially weaker. **B.** In spite of the competitive interactions
between the two shrub species, their shared tendency to occur in locations
without trees makes their occurrence vectors positively correlated (circled).
**C.** Controlling for the tree species' presence with a conditional method
such as a partial covariance or a Markov network allows us to correctly
identify the negative interaction between these two species (circled).

**Figure 2.**
**A.** A small Markov network with two species.  The depicted abiotic
environment favors the occurrence of both species ($\alpha >0$), particularly
species 2 ($\alpha_2 > \alpha_1$). The  negative $\beta$ coefficient linking
these two species implies that they co-occur less than expected under
independence.  **B.** Relative probabilities of all four possible presence-absence
combinations for Species 1 and Species 2. The exponent includes
$\alpha_1$ whenever Species 1 is present ($y_1 = 1$), but not when it is
absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
species $2$ is present ($y_2 = 1$), and $\beta$ only when both are present
($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four relative
probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** Using the
probabilities, we can find the expected frequencies of all possible co-occurrence
 patterns between the two species of interest. **D.** If $\beta$
equaled zero (e.g. if the species no longer competed for the same resources),
then the reduction in competition would allow each species to increase its
occurrence rate and the deficit of co-occurrences would be eliminated.

**Figure 3.**
Proportion of variance in interaction coefficients explained by each method
across varying numbers of sampled locations.
